Microlensing as a possible probe of event-horizon structure in quasars

### 2017-09-11 paper1/msold3.tex
edits

In the present study we analyse three types of sources with different surface brightness: 
\begin{enumerate}
 \renewcommand{\theenumi}{(\arabic{enumi})}
  \item a rotationally symmetric source with a bivariate Gaussian surface brightness distribution,
  \item a disk source with constant surface brightness distribution,
  \item a crescent shaped source with constant surface brightness distribution.
\end{enumerate}
The first two sources are the typical choices used in the literature to describe the luminous parts of a quasar. 
The third one is a recently proposed variant \citep{2013MNRAS.434..765K}.

\subsection{Rotationally symmetric source with a bivariate Gaussian surface brightness distribution}\label{subsec:gaussian}

A symmetric 2D Gaussian can be described mathematically as:

@@

Although such a definition for a source would have non-zero surface/linear brightness for any coordinate $p,q$, the amount of light received by a detector from outside a $3 \sigma$ disk centered at $p_s, q_s$ 
would be insignificant. For a Gaussian distributed surface brightness source the half-light radius is directly proportional to the parameter $\sigma$ according to the equation:
\begin{equation}
r_{1/2} = \sqrt{\ln4}\,\sigma.
\end{equation}

@@

One can construct mathematically a disk source with constant surface brightness and radius $R$ using a stepfunction:
\begin{equation}
 S_{2D}^D(p-p_s, q-q_s) = \frac{S_0^D}{\pi R^2} \Theta \left( R^2 - \left( p-p_s \right)^2 - \left( q-q_s \right)^2 \right).
\end{equation}
By integrating over $q$ coordinate the linear brightness function is obtained:

@@

\subsection{Lightcurve of the Gaussian source}

The amount of light received by an observer from a source 
with a Gaussian distributed brightness with $\sigma$ and total flux $S_0^G$ in the absence of any gravitational lensing is:
\begin{equation}
 F^G(t) = \int_{-\infty}^\infty  \left( \mu_0 + \frac{C_0}{\sqrt{p}} \Theta \left( p \right) \right) \left( \frac{S_0^G}{\sqrt{2 \pi} \sigma} e^{-\frac{(p-p_s(t))^2}{2 \sigma^2}} \right) \mathrm{d}p.
\end{equation}
which can be simplified to:
\begin{equation}
 F^G(t) = \mu_0 S_0^G + \frac{C_0 S_0^G}{\sqrt{2\pi} \sigma} \int_{0}^\infty \frac{e^{-\frac{(p-p_s(t))^2}{2 \sigma^2}}}{\sqrt{p}} \mathrm{d}p.
\end{equation}

\begin{figure}
\centering
        \includegraphics[width = 0.48\textwidth]{figures/ch_points.eps}
        \includegraphics[width = 0.48\textwidth]{figures/ch_instances.eps}
\caption{\label{fig:char_points} The characteristic points and
  instances of a crescent source's one dimensional profile and
  microlensing lightcurve.  In the top panel a typical 1D profile of a
  crescent source is presented.  There are two light gray regions that
  mark the regions of the bright disk that are not overlapped by the
  dark disk. The dark gray band marks the region where the two disks
  overlap. The points $p_1$ and $p_4$ mark the start and end of the
  bright disk in the 1D representation. Analogously, $p_2$ and $p_3$
  mark the start and end of the dark disk.  The bottom panel
  represents the associated lightcurve with the points $t_i$ marking
  the moment when point $p_i$ overlaps with the caustic.  The gray
  shades are analogous to the ones in the top panel.  }
\end{figure}

@@

Analogous to the Gaussian shaped source, the disk source with uniform brightness, radius $R$ and 
unmagnified flux $S_0^D$ has a lightcurve described by the equation:

@@

Figure \ref{fig:lightcurve_gauss} reveals that the 
lightcurve of a crescent source has more
visible features than the other two light-curves corresponding to the
disc and Gaussian shape. There are three regions where $S_{1D}$ and by
inheritance, the lightcurve has distinguishable behaviour. The first
region that would be recorded on a lightcurve plot represents the
period of time when the bright disk begins to be overlapped by the
caustic and stops when just before the dark disk reaches the
caustic. During this period of time, the flux of light from the source
is increasingly magnified. The second period starts and ends with the
overlapping of the dark disc. As a boundary of the two regions, there
is a distinguishable point where the slope of the magnification is
drastically changed. This apparent discontinuity in the first
derivative of the magnification function is caused by the caustic
amplification of the sudden drop in the $S_{1d}$ function. During the
respective period, the magnification growth slows or even reverses
and later starts to grow faster again as
the dark disc ends its overlap with the caustic. At the point where
the dark disc clears the caustic, the growth of the magnification is
infinite, which appears as a saddle point on the lightcurve. Next, the
final period corresponds to the case when the dark disc has cleared
the caustic and the bright disc continues to overlap with the
caustic. During this period, the lightcurve reaches a peak that for
most of the parameter space is global and for the rest of the
parameter space local.  At the end of the period, the growth of the
magnification is negative infinite. The respective point appears as a
second saddle point on the lightcurve. Past this point the
magnification of any finite source will decay in roughly the same
manner.

@@

\begin{figure}
\centering
    \includegraphics[width = 0.48\textwidth]{figures/M87_shape.eps}
        \includegraphics[width = 0.48\textwidth]{figures/M87_lc.eps}
\caption{\label{fig:M87_plots} The one dimensional profile obtained
  from the numerical integration along the ordonata axis of a
  simulated source image is presented in the upper panel. The source
  image was presented in Dexter et al (2012) in the top right-most
  panel of Figure~5 (of the respective paper) as a result of the
  simulations presented in McKinney \& Blandford (2009). The
  corresponding lightcurve associated to the 1D profile is presented
  in the lower panel. The points $\rm p_2$ and $\rm p_3$ mark the
  start and end of the equivalent dark disk and are associated with
  the moments of their overlap with the caustic: $t_2$ and $t_3$;
  $h_{w}$ denotes the maximum width of the crescent.}
\end{figure}

@@

Having used the simple model of an ideal fold to gain insight, we now
consider the question of whether the characteristics of a crescent
source could be discerned in real data having noise and systematic
error as well.  Do do so, we first generate a microlensing lightcurve
using a widely-used microlensing code
\citep{1990PhDT.......180W,1999A&A...346L...5W,1999JCoAM.109..353W} to
which we have added a small extension to provide the option of
crescent sources.  We then fit model or template lightcurves ---which
are generated with varied source and other parameters--- to the mock
lightcurve.  The basic technique is the same as in previous studies
\citep[e.g.,][]{2010ApJ...712..658P} though the details are different.

@@

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{figures/cc_backward.eps}
\caption{\label{fig:cc_backward} Like Figure~\ref{fig:cc_forward},
  except that the fit is derived from a slightly different caustic
  than the one used to generate the mock data.  The curve has
  $\chi^2_{(244)}=284, P=0.01$ indicating a marginally acceptable
  fit.}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{figures/gc_forward.eps}
\caption{\label{fig:gc_forward} Like Figures~\ref{fig:cc_forward} and
  \ref{fig:cc_backward}, but showing the best fit to a Gaussian
  source.  The histogram shows the posterior distribution of $r_{1/2}$
  (there is no $R_n$).  It appears to show $r_{1/2}$ as accurately
  recovered, but is not to be trusted, because the fit has
  $\chi^2_{(247)}=344, P=3\times10^{-6}$ indicating rejection of the
  Gaussian source.  Looking closely, the curve is a little too shallow
  near $t=3$, and it is unable to fit the second sub-peak near $t=5$.}
\end{figure}

@@

Nevertheless, the respective displacement is relevant for the calculation of
of the half-light radius.  Its qualitative effect is to 
decrease the value of the
half-light radius when the absolute value of the parameter is
increased.  In contrast to the simpler source shapes discussed before the
crescent has two additional relevant parameters: the radius of the dark 
disk ($R_n$) and the center displacement in the direction
 perpendicular to the caustic $a$.

Figure \ref{fig:lightcurve_gauss} reveals that the lightcurve of a crescent source has more
visible features than the other two light-curves corresponding to the
disc and Gaussian shape. The parameters $R_n$ and $a$ have strong
influences on the shape of the microlensed lightcurve as can be seen
in Figures~\ref{fig:a_var}, \ref{fig:lightcurve_crescent_back} and
\ref{fig:lightcurve_crescent}.  Moreover, the one-dimensional
source profile corresponding to the direction perpendicular to the
caustic reveal four characteristic points. The overlap of each of
these points with the caustic leaves visible features on the
lightcurve at the corresponding instances of time. In time order,
the instances correspond to the start of the overlap between the
caustic and the bright disk, the start of the overlap between the
caustic and the dark disk, the end of the overlap between the caustic
and dark disk and finally the end of the overlap between the caustic
and the bright disk.

@@

Furthermore, a simulated image of M87 presented in
\cite{2012MNRAS.421.1517D} has been microlensed
(Figure~\ref{fig:M87_plots}).  On the resulting lightcurve the
instances corresponding to the start and end of the black hole shadow
and caustic overlap were distinguishable.

@@

The previously mentioned abstract parameters can be related to
physical quantities specific to the central region of a quasar.  As
such the luminous region would correspond to the bright accretion disc
that surrounds the black hole. The later's gravity would cause a
shadow in the bright region limited by the extent of the event horizon
of the black hole. Therefore, the radius of the bright disk would
provide an estimate of the size of the accretion disk and the radius
of the dark disc would provide and estimate of the gravitationally
magnified Schwarzschild radius of the black hole $R_{S}^{magnified} =
\Delta t_{dark} \cdot v_p$.  By $R_{S}^{magnified}$ we refer to the
apparent Schwarzschild radius which is larger than the real value at
large distances due to the black hole's own gravity.  Moreover, the
gravitationally magnified value of the Schwarzschild radius is a
monotonic function of the black hole's mass. Therefore, it can be used
to estimate the mass of the black hole if it was not rotating. In the
previous expression, the $\Delta t_{dark}$ denotes the period of
overlap between the black hole shadow and fold caustic, while $v_p$
denotes the component of the relative velocity of the source and fold
which is perpendicular to the caustic.  The respective velocity is an
unknown, though it can be constrained on a case by case basis to an
order of magnitude or even better. This would require the study of the
dynamics of the stellar structure which contains the gravitational
lens and the bulk transverse velocity of the galaxy with respect to
the observer.  A better estimate of the relative velocity would
facilitate a better estimate of the effective non-rotating black hole
mass associated with the black hole shadow.

@@

\def\apj{ApJ}
\def\apjl{ApJL}
\def\aj{AJ}
\def\mnras{MNRAS}
\def\aap{A\&A}
\def\nat{nature}
\def\araa{ARAA}
\def\pasa{PASA}
\bibliography{heap}

>>>>

In the present study we analyse three types of sources with different surface brightness: 
\begin{enumerate}
 \renewcommand{\theenumi}{(\arabic{enumi})}
  \item a rotationally symmetric source with a bivariate gaussian surface brightness distribution,
  \item a disk source with constant surface brightness distribution,
  \item a crescent shaped source with constant surface brightness distribution.
\end{enumerate}
The first two sources are the typical choices used in the literature to describe the luminous parts of a quasar. 
The third one is a recently proposed variant \citep{2013MNRAS.434..765K}.

\subsection{Rotationally symmetric source with a bivariate gaussian surface brightness distribution}\label{subsec:gaussian}

A symmetric 2D gaussian can be described mathematically as:

@@

Although such a definition for a source would have non-zero surface/linear brightness for any coordinate $p,q$, the amount of light received by a detector from outside a $3 \sigma$ disk centered at $p_s, q_s$ 
would be insignificant. For a gaussian distributed surface brightness source the half-light radius is directly proportional to the parameter $\sigma$ according to the equation:
\begin{equation}
r_{1/2} = \sqrt{\ln4}\,\sigma.
\end{equation}

@@

One can construct mathematically a disk source with constant surface-brightness and radius $R$ using a stepfunction:
\begin{equation}
 S_{2D}^D(p-p_s, q-q_s) = \frac{S_0^D}{\pi R^2} \Theta \left( R^2 - \left( p-p_s \right)^2 - \left( q-q_s \right)^2 \right).
\end{equation}
By integrating over $q$ coordinate the linear brightness function is obtained:

@@

\subsection{Lightcurve of the gaussian source}

The amount of light received by an observer from a source 
with a gaussian distributed brightness with $\sigma$ and total flux $S_0^G$ in the absence of any gravitational lensing is:
\begin{equation}
 F^G(t) = \int_{-\infty}^\infty  \left( \mu_0 + \frac{C_0}{\sqrt{p}} \Theta \left( p \right) \right) \left( \frac{S_0^G}{\sqrt{2 \pi} \sigma} e^{-\frac{(p-p_s(t))^2}{2 \sigma^2}} \right) \mathrm{d}p.
\end{equation}
which can be simplified to:
\begin{equation}
 F^G(t) = \mu_0 S_0^G + \frac{C_0 S_0^G}{\sqrt{2\pi} \sigma} \int_{0}^\infty \frac{e^{-\frac{(p-p_s(t))^2}{2 \sigma^2}}}{\sqrt{p}} \mathrm{d}p.
\end{equation}

\begin{figure}
\centering
        \includegraphics[width = 0.48\textwidth]{figures/ch_points.eps}
        \includegraphics[width = 0.48\textwidth]{figures/ch_instances.eps}
\caption{\label{fig:char_points} The characteristic points and instances
of a crescent source's one dimensional profile and microlensing lightcurve.
In the top panel a typical 1D profile  of a crescent source is presented.
There are two light gray regions that mark the regions of the bright disk
that are not overlapped by the dark disk. The dark gray band marks the
region where the two disks overlap. $p_1$ and $p_4$ mark the start and
end of the bright disk in the
1D representation. Analogous, $p_2$ and $p_3$ mark the start and
end of the dark disk.
The bottom panel represents the associated lightcurve with the points
$t_i$ marking the moment when point $p_i$ overlaps with the caustic.
The gray shades are analogous to the ones in the top panel.
 }
\end{figure}

@@

Analogous to the gaussian shaped source, the disk source with uniform brightness, radius $R$ and 
unmagnified flux $S_0^D$ has a lightcurve described by the equation:

@@

Figure \ref{fig:lightcurve_gauss} reveals that the 
lightcurve of a crescent source has more
visible features than the other two light-curves corresponding to the
disc and gaussian shape. There are three regions where $S_{1D}$ and by
inheritance, the lightcurve has distinguishable behaviour. The first
region that would be recorded on a lightcurve plot represents the
period of time when the bright disk begins to be overlapped by the
caustic and stops when just before the dark disk reaches the
caustic. During this period of time, the flux of light from the source
is increasingly magnified. The second period starts and ends with the
overlapping of the dark disc. As a boundary of the two regions, there
is a distinguishable point where the slope of the magnification is
drastically changed. This apparent discontinuity in the first
derivative of the magnification function is caused by the caustic
amplification of the sudden drop in the $S_{1d}$ function. During the
respective period, the magnification growth slows or even reverses
and later starts to grow faster again as
the dark disc ends its overlap with the caustic. At the point where
the dark disc clears the caustic, the growth of the magnification is
infinite, which appears as a saddle point on the lightcurve. Next, the
final period corresponds to the case when the dark disc has cleared
the caustic and the bright disc continues to overlap with the
caustic. During this period, the lightcurve reaches a peak that for
most of the parameter space is global and for the rest of the
parameter space local.  At the end of the period, the growth of the
magnification is negative infinite. The respective point appears as a
second saddle point on the lightcurve. Past this point the
magnification of any finite source will decay in roughly the same
manner.

@@

\begin{figure}
\centering
    \includegraphics[width = 0.48\textwidth]{figures/M87_shape.eps}
        \includegraphics[width = 0.48\textwidth]{figures/M87_lc.eps}
\caption{\label{fig:M87_plots} The one dimensional profile obtained
  from the numerical integration along the ordonata axis of a simulated source
  image is presented in the upper panel. The source image was presented in 
 Dexter et al 2012 in the top right-most panel of figure 5 
(of the respective paper) as a result of the simulations 
presented in McKinney \& Blandford 2009. The corresponding
  lightcurve associated to the 1D profile is presented in the lower
  panel. The points $\rm p_2$ and $\rm p_3$ mark the start and 
  end of the equivalent
  dark disk and are associated with the moments of their overlap with
  the caustic: $\rm t_2$ and $\rm t_3$. $\rm h_{w}$ denotes the maximum width of the crescent.}
\end{figure}

@@

Having used the simple model of an ideal fold to gain insight, we now
consider the question of whether the characteristics of a crescent
source could be discerned in real data having noise and systematic
error as well.  Our method is to generate microlensing lightcurves
using a widely-used microlensing code
\citep{1990PhDT.......180W,1999A&A...346L...5W,1999JCoAM.109..353W} to
which we have added a small extension to provide the option of
crescent sources, and then fit these mock lightcurves to source models
using Markov-chain Monte Carlo (MCMC).  The basic technique is the same
as in previous studies \citep[e.g.,][]{2010ApJ...712..658P} though
the details are different.

@@

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{figures/cc_backward.eps}
\caption{\label{fig:cc_backward} Like Figure~\ref{fig:cc_backward},
  except that the fit uses path~II whereas the mock data still come
  from path~I. The curve has $\chi^2_{(244)}=284, P=0.01$ indicating a
  marginally acceptable fit.}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{figures/gc_forward.eps}
\caption{\label{fig:gc_forward} Like Figure~\ref{fig:cc_backward}, but
  showing the best fit to a Gaussian source.  The histogram shows the
  posterior distribution of $r_{1/2}$ (there is no $R_n$).  It appears
  to show $r_{1/2}$ as accurately recovered, but is not to be trusted,
  because the fit has $\chi^2_{(247)}=344, P=3\times10^{-6}$ indicating
  rejection of the Gaussian source.  Looking closely, the curve is a
  little too shallow near $t=3$, and it is unable to fit the second
  sub-peak near $t=5$.}
\end{figure}

@@

Nevertheless, the respective displacement is relevant for the calculation of
of the half-light radius.  It's qualitative effect is to 
decrease the value of the
half-light radius when the absolute value of the parameter is
increased.  In contrast to the simpler source shapes discussed before the
crescent has two additional relevant parameters: the radius of the dark 
disk ($R_n$) and the center displacement in the direction
 perpendicular to the caustic $a$.

Figure \ref{fig:lightcurve_gauss} reveals that the lightcurve of a crescent source has more
visible features than the other two light-curves corresponding to the
disc and Gaussian shape. The parameters $R_n$ and $a$ have strong
influences on the shape of the microlensed lightcurve as can be seen
in Figures~\ref{fig:a_var}, \ref{fig:lightcurve_crescent} and
\ref{fig:lightcurve_crescent_back}.  Moreover, the one-dimensional
source profile corresponding to the direction perpendicular to the
caustic reveal four characteristic points. The overlap of each of
these points with the caustic leaves visible features on the
lightcurve at the corresponding instances of time. In timely order,
the instances correspond to the start of the overlap between the
caustic and the bright disk, the start of the overlap between the
caustic and the dark disk, the end of the overlap between the caustic
and dark disk and finally the end of the overlap between the caustic
and the bright disk.

@@

Furthermore, a simulated image of M87 presented in
 \citep{2012MNRAS.421.1517D} has been microlensed (figure \ref{fig:M87_plots}). 
On the resulting lightcurve the instances 
corresponding to the start and end of the black hole 
shadow and caustic overlap were distinguishable.

@@

The previously mentioned abstract parameters can be related to
physical quantities specific to the central region of a quasar.  As
such the luminous region would correspond to the bright accretion disc
that surrounds the black hole. The later's gravity would cause a
shadow in the bright region limited by the extent of the event horizon
of the black hole. Therefore, the radius of the bright disk would
provide an estimate of the size of the accretion disk and the radius
of the dark disc would provide and estimate of the gravitationally
magnified Schwarzschild radius of the black hole $R_{S}^{magnified} =
\Delta t_{dark} \cdot v_p$.  By $R_{S}^{magnified}$ we refer to the
apparent Schwarzschild radius which is larger than the real value at
large distances due to the black hole's own gravity.  Moreover, the
gravitationally magnified value of the Schwarzschild radius is a
monotonic function of the black hole's mass. Therefore, it can be used
to estimate the mass of the black hole if it was not rotating. In the
previous expression, the $\Delta t_{dark}$ denotes the period of
overlap between the black hole shadow and fold caustic.  $v_p$ denotes
the component of the relative velocity of the source and fold which is
perpendicular to the caustic.  The respective velocity is an unknown,
though it can be constrained on a case by case basis to an order of
magnitude or even better. This would require the study of the dynamics
of the stellar structure which contains the gravitational lens and the bulk
transverse velocity of the galaxy with respect to the observer.  A
better estimate of the relative velocity would facilitate a better
estimate of the effective non-rotating black hole mass associated with
the black hole shadow.

@@

\def\apj{ApJ}
\def\apjl{ApJL}
\def\aj{AJ}
\def\mnras{MNRAS}
\def\aap{A\&A}
\def\nat{Nature}
\def\araa{ARAA}
\def\pasa{PASA}
\bibliography{heap}
